---
title: "iCARE model validation — reproduce in R (Quarto + OJS)"
format:
  html:
    embed-resources: true
    toc: true
engine: knitr
---

R users can drive the same `wasm-icare` engine the app uses from a Quarto notebook: an **R chunk**
reads the input files and hands them to an **`{ojs}` (Observable JavaScript) cell**, which loads
`wasm-icare` from a CDN and runs the validation client-side in your browser. Nothing is uploaded.

> **How data crosses from R to OJS.** Quarto's `ojs_define()` serializes an R value to JSON and exposes
> it to `{ojs}` cells by name. `wasm-icare`'s model inputs must be *file-like* (a path/URL/`Blob`) — the
> incidence tables, reference dataset, covariate profile, and the log-OR JSON are **not** accepted as
> in-memory JS arrays/objects (an inline formula *string* is the one exception). So we serialize each
> file's **raw text** from R and rebuild a `Blob` from it in OJS. This is byte-faithful — it also
> preserves the `Inf` in `time_of_onset` (censored rows), which a numeric JSON round-trip would mangle.

## 1 · Load the inputs and serialize them to OJS (R)

```{r}
#| label: load-inputs
library(readr)

# EDIT these paths to point at your own files. They default to this repo's bundled iCARE-Lit ge50
# example, so the notebook renders as-is from the repo root.
dir <- "../../public/examples/icare-lit"

# Read each file as raw text and expose it to OJS. `read_file()` keeps the bytes exactly (incl. "Inf").
ojs_define(
  studyText     = read_file(file.path(dir, "icare_lit_validation_study.csv")),
  diseaseText   = read_file(file.path(dir, "age_specific_breast_cancer_incidence_rates.csv")),
  competingText = read_file(file.path(dir, "age_specific_all_cause_mortality_rates.csv")),
  formulaText   = read_file(file.path(dir, "model_formula_ge50.txt")),
  logOrText     = read_file(file.path(dir, "model_log_odds_ratios_ge50.json")),
  referenceText = read_file(file.path(dir, "reference_covariate_data_ge50.csv")),
  profileText   = read_file(file.path(dir, "icare_lit_validation_covariates.csv"))
)
```

### Inspecting a data frame across the R → OJS boundary (optional)

`ojs_define()` can also serialize a *data frame*. Quarto delivers it **column-oriented** to OJS; the
built-in `transpose()` turns it into an array of row objects — the idiomatic pattern for driving OJS
tables/plots from R. (Here it is just for a preview; the validation above uses the raw-text `Blob`s.)

```{r}
#| label: preview-df
study_df <- readr::read_csv(file.path(dir, "icare_lit_validation_study.csv"), show_col_types = FALSE)
ojs_define(studyPreview = head(study_df, 8))
```

```{ojs}
//| label: preview-table
Inputs.table(transpose(studyPreview))
```

## 2 · Run the validation (OJS, in your browser)

Loading the engine pulls Pyodide + `pyicare` from the jsDelivr CDN the first time (~40 MB, ~20–30 s;
needs internet). For a fully offline notebook, self-host a Pyodide mirror and pass
`indexURL`/`pyicareWheelUrl` to `loadICARE` (see the `wasm-icare` README, "Self-hosting / offline").

```{ojs}
//| label: load-engine
icare = (await import("https://esm.sh/wasm-icare@2")).loadICARE()
```

```{ojs}
//| label: validate
result = {
  const i = await icare;
  const v = await i.validateAbsoluteRiskModel({
    // Rebuild file-like Blobs from the raw text serialized by R above.
    studyData: new Blob([studyText]),
    predictedRiskInterval: "total-followup",
    icareModelParameters: {
      modelDiseaseIncidenceRates: new Blob([diseaseText]),
      modelCompetingIncidenceRates: new Blob([competingText]),
      modelCovariateFormula: formulaText,          // an inline formula string is accepted
      modelLogRelativeRisk: new Blob([logOrText]), // ...but the log-OR must be a Blob of its JSON
      modelReferenceDataset: new Blob([referenceText]),
      applyCovariateProfile: new Blob([profileText]),
    },
    numberOfPercentiles: 10,
    seed: 50,
    datasetName: "iCARE-Lit ge50",
    modelName: "iCARE-Lit",
  });
  window.__ojsResult = v; // exposed for automated verification
  return v;
}
```

## 3 · Results (OJS)

```{ojs}
//| label: metrics
md`
| Metric | Value |
|---|---|
| **AUC** | ${result.auc.auc.toFixed(4)} [${result.auc.lowerCi.toFixed(4)}, ${result.auc.upperCi.toFixed(4)}] |
| **E/O ratio** | ${result.expectedByObservedRatio.ratio.toFixed(4)} |
| **Hosmer-Lemeshow χ²** | ${result.calibration.absoluteRisk.statistic.chiSquare.toFixed(4)} (df ${result.calibration.absoluteRisk.parameter.degreesOfFreedom}) |

_Expected for the bundled iCARE-Lit ge50 example: AUC ≈ 0.6341, E/O ≈ 1.0275, HL χ² ≈ 23.17 (df 10)._
`
```

```{ojs}
//| label: calibration-table
Inputs.table(transpose(result.categorySpecificCalibration.columns), {
  columns: ["category", "observed_absolute_risk", "predicted_absolute_risk", "expected_by_observed_ratio"],
  header: {
    category: "Decile",
    observed_absolute_risk: "Observed",
    predicted_absolute_risk: "Predicted",
    expected_by_observed_ratio: "E/O",
  },
})
```
